---
title: "Introduction to slitherlinkshiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to slitherlinkshiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is Slitherlink?

Slitherlink is a logic puzzle played on a rectangular grid of dots.
The goal is to draw segments between adjacent dots to form a **single
closed loop** — no branches, no crossings, no loose ends.

Numbers inside cells indicate exactly how many of the cell's four
edges belong to the loop.  Cells without a number are unconstrained.

```
+   +   +   +
  2   1   2
+   +   +   +
  1   0   1
+   +   +   +
  2   1   2
+   +   +   +
```

The unique solution for the 3×3 puzzle above is the outer rectangle.

---

## Grid representation

The package stores a puzzle as a `slitherlink_grid` object with three
components:

| Field    | Dimensions        | Meaning                      |
|----------|-------------------|------------------------------|
| `clues`  | n × m             | Clue values (NA = no clue)   |
| `seg_h`  | (n+1) × m         | Horizontal segment states    |
| `seg_v`  | n × (m+1)         | Vertical segment states      |

Each segment is one of three integer values:

| Value | Meaning  |
|-------|----------|
|  `0`  | Empty    |
|  `1`  | Drawn    |
| `-1`  | Crossed  |

---

## Creating a grid

Use `init_grid()` to build a grid from a clue matrix.
`NA` marks cells with no constraint.

```{r create}
library(slitherlinkshiny)

clues <- matrix(
  c(2, 1, 2,
    1, 0, 1,
    2, 1, 2),
  nrow = 3, byrow = TRUE
)
g <- init_grid(clues)
print(g)
```

---

## Predefined puzzles

The package ships with seven ready-to-play puzzles spanning three difficulty levels.

```{r puzzles}
list_puzzles()
```

Load one with `get_puzzle()`:

```{r load_puzzle}
g <- get_puzzle("medium_4x4")
print(g)
```

---

## Toggling segments

`toggle_segment()` cycles a segment through the three states:
`0 → 1 → -1 → 0`.

- `type = "h"` targets a horizontal segment at row `i`, column `j`
  (row index counts from top; `i` ranges from 1 to n+1).
- `type = "v"` targets a vertical segment (`i` from 1 to n,
  `j` from 1 to m+1).

```{r toggle}
g <- get_puzzle("easy_3x3")

# Draw the top edge of cell (1,1): horizontal segment h[1,1]
g <- toggle_segment(g, type = "h", i = 1, j = 1)
g$seg_h[1, 1]   # now 1 (drawn)

# Draw it again to cross it
g <- toggle_segment(g, type = "h", i = 1, j = 1)
g$seg_h[1, 1]   # now -1 (crossed)

# Draw it once more to reset
g <- toggle_segment(g, type = "h", i = 1, j = 1)
g$seg_h[1, 1]   # back to 0 (empty)
```

---

## Validating the state

Three functions check the current state of the grid.

### `check_clues()`

Verifies that no clue constraint is violated.

- `strict = FALSE` (default for in-progress play): drawn edges must
  not *exceed* the clue.
- `strict = TRUE`: drawn edges must *exactly equal* the clue.

```{r check_clues}
g <- get_puzzle("easy_3x3")
check_clues(g, strict = FALSE)   # TRUE  — nothing drawn yet
check_clues(g, strict = TRUE)    # FALSE — no clue is exactly matched
```

### `check_loop()`

Returns `TRUE` if the drawn segments form a single valid closed loop
(every active node has degree exactly 2, and all active nodes are
connected).

```{r check_loop}
check_loop(g)   # FALSE — no segments drawn
```

### `is_solved()`

A puzzle is solved when both `check_clues(strict = TRUE)` and
`check_loop()` return `TRUE`.

```{r is_solved}
is_solved(g)   # FALSE
```

---

## Solving a puzzle automatically

`solve_grid()` uses constraint propagation followed by backtracking
to find the unique solution.

```{r solve}
g   <- get_puzzle("easy_3x3")
sol <- solve_grid(g)
print(sol)
is_solved(sol)
```

The solver works on all built-in puzzles:

```{r solve_all}
for (nm in list_puzzles()$name) {
  sol <- solve_grid(get_puzzle(nm))
  cat(nm, "→ solved:", is_solved(sol), "\n")
}
```

If the puzzle has no valid solution, `solve_grid()` returns `NULL`:

```{r solve_null}
impossible <- init_grid(matrix(2L, 1, 1))  # clue 2 in a 1x1 grid: impossible
is.null(solve_grid(impossible))
```

---

## How the solver works

`solve_grid()` combines two classical techniques from constraint programming.

### Constraint propagation

Before any branching, the solver applies deduction rules exhaustively until
no further progress can be made (fixpoint).

**Cell rules** — for every cell with a clue $c$, let $d$ = drawn edges and
$f$ = free edges:

- If $d = c$: all $f$ remaining free edges must be crossed (the clue is
  already satisfied).
- If $d + f = c$: all $f$ free edges must be drawn (every remaining edge is
  needed).

**Node rules** — for every grid intersection:

- A node with two drawn edges already has its required degree; all remaining
  free edges at that node are crossed.
- A node with one drawn edge and exactly one free edge left must use that
  edge (a valid loop requires degree 0 or 2 at every node).

The engine also detects three types of contradiction that prove a partial
assignment cannot lead to a solution: a clue is exceeded, a node reaches
degree > 2, or a sub-loop closes prematurely (a connected component of
drawn edges already forms a cycle but not all drawn edges are part of it).

Propagation repeats until neither `seg_h` nor `seg_v` changes, or a
contradiction is found.

### Backtracking with MRV heuristic

When propagation stalls on an undecided segment, the solver branches: it
tries drawing the segment, then crossing it, recursively.

Rather than picking the first undecided segment, it uses the **Minimum
Remaining Values (MRV)** heuristic: the chosen segment is adjacent to the
cell with the smallest *slack* (free edges minus remaining edges still
needed). Branching on the most constrained cell reduces the search tree
significantly.

### Uniqueness verification

`random_puzzle()` calls an internal `count_solutions()` function that
counts distinct solutions up to a given limit and stops as soon as the
limit is reached.  Any candidate grid that admits two or more solutions is
discarded, and the generation loop retries with a new random region.  This
guarantees that every puzzle returned by `random_puzzle()` has exactly one
solution.

---

## Interactive Shiny app

The package includes a full interactive web application.
Launch it from the R console:

```r
run_app()
```

Features:

- **Select puzzle** — choose from the built-in library.
- **New Game / Reset** — load or restart the selected puzzle.
- **Click segments** — click near any edge to cycle its state
  (empty → drawn → crossed → empty).
- **Solve** — fill in the complete solution instantly.
- **Status indicator** — shows *In progress*, *Constraint violated*,
  or *Puzzle solved!* in real time.
